Automated assessment of residual plots with computer vision models

### 2024-04-29 paper/paper.tex
update diagram
revise up to model arch

\usepackage{lscape}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\def\tightlist{}
\usepackage{setspace}
\doublespacing

@@

Generating a residual plot in most statistical software is often as
straightforward as executing a line of code or clicking a button.
However, accurately interpreting a residual plot can be challenging.
Consider Figure \ref{fig:false-finding} as an example, the residuals
display a triangular shape pointing to the left. While this might
suggest heteroskedasticity, it is important to avoid over-interpreting
this visual pattern. In this case, the fitted regression model is
correctly specified, and the triangular shape is actually a result of
the skewed distribution of the predictors, rather than indicating a flaw
in the model.

A residual plot can exhibit various visual features, but it is crucial
to recognize that some may arise from the characteristics of predictors
and the inherent randomness of the error, rather than indicating a
violation of model assumptions \citep{li2023plot}. The concept of visual
inference, as proposed by \citet{buja2009statistical}, provides an
inferential framework to assess whether residual plots indeed contain
visual patterns inconsistent with the model assumptions. The fundamental
idea involves testing whether the actual residual plot visually differs
significantly from null plots, where null plots are plotted with
residuals generated from the residual rotation distribution
\citep{langsrud2005rotation}, which is a distribution consistent with
the null hypothesis \(H_0\) that the linear regression model is
correctly specified. Typically, the visual test is accomplished through
the lineup protocol, where the real residual plot is embedded within a
lineup alongside several null plots. If the real residual plot can be
distinguished from the lineup, it provides evidence for rejecting
\(H_0\).

@@

The utilization of computers to interpret data plots has a rich history,
with early efforts such as ``Scagnostics'' by \citet{tukey1985computer},
focusing on scatter plot diagnostics. \citet{wilkinson2005graph}
expanded on this work, introducing graph theoretic scagnostics, which
encompassed computable measures applied to planar proximity graphs.
These measures, including, but not limited to, ``Outlying'', ``Skinny'',
``Stringy'', ``Straight'', ``Monotonic'', ``Skewed'', ``Clumpy'', and
``Striated'' aimed to characterize outliers, shape, density, trend,
coherence and other characteristics of the data. While this approach has
been inspiring, there is a recognition \citep{buja2009statistical} that
it may not capture all the necessary visual features that differentiate
actual residual plots from null plots. A more promising alternative
entails enabling machines to learn the function for extracting visual
features from residual plots. Essentially, this means empowering
computers to discern the crucial visual features for residual
diagnostics and determining the method to extract them.

@@

\subsection{Different possible configurations of the model
formula}\label{different-possible-configurations-of-the-model-formula}

There are various configurations of the computer vision model that can
be used to assess residual plots. We discuss these configurations below
based on two key components of the model formula: the input and the
output format.

\subsubsection{Input formats}\label{input-formats}

@@

Providing multiple residual plots to the model, such as a pair of plots,
a triplet or a lineup is also a possible option.
\citet{chopra2005learning} have shown that computer vision models
designed for image comparison can assess whether a pair of images are
similar or dissimilar. Applied to our specific problem, we can define
null plots of a fitted regression model to be similar to each other,
while considering actual residual plots to be distinct from null plots
of any fitted regression model. A triplet constitutes a set of three
images, denoted as \(image_1\), \(image_2\) and \(image_3\). It is often
used to predict whether \(image_2\) or \(image_3\) is more similar to
\(image_1\), proving particularly useful for establishing rankings
between samples. For this setup, we can apply the same criteria to
define similarity between images. However, it is important to note that
these two approaches usually require additional considerations regarding
the loss function and, at times, non-standard training processes due to
shared weights between different convolutional blocks.

@@

We did not explore all the mentioned input formats due the considerable
costs associated with data preparation and model training. Considering
the implementation cost and the interpretability of the model, we
settled on the single residual plot input format.

\subsubsection{Output formats}\label{output-formats}

@@

\subsection{Distance from the good residual
plots}\label{distance-from-the-good-residual-plots}

@@

\subsubsection{Residual distribution}\label{residual-distribution}

The distance measure proposed in this study takes into account the fact
that we tried to measure how different a residual plot is from a good
residual plot, or in other words, how different a given fitted
regression model is from a correctly specified model. For the classical
normal linear regression model, residuals \(\boldsymbol{e}\) are derived
from the fitted values \(\hat{\boldsymbol{y}}\) and observed values
\(\boldsymbol{y}\). Suppose the data generating process is known and the
regression model is correctly specified, by the Frisch-Waugh-Lowell
theorem \citep{frisch1933partial}, residuals \(\boldsymbol{e}\) can also
be treated as random variables and written as a linear transformation of
the error \(\boldsymbol{\varepsilon}\) formulated as
\(\boldsymbol{e} = \boldsymbol{R}\boldsymbol{\varepsilon}\), where
\(\boldsymbol{R}=\boldsymbol{I}_n -\boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'\)
is the residual operator and an idempotent matrix, \(\boldsymbol{X}\) is
the design matrix, \(\boldsymbol{I}_n\) is a \(n\) by \(n\) identity
matrix, and \(n\) is the number of observations.

@@

In ordinary least square, the minimization of the sum of square
residuals implies \(\sum_{i=1}^{n} e_i = 0\), making any residual value
to be a linear combination of the remaining \(n - 1\) residuals. This
effectively means \(\text{rank}(\boldsymbol{R}) = n - 1 < n\) and \(Q\)
is a degenerate multivariate distribution. To capture the
characteristics of \(Q\), such as moments, we can simulate a large
numbers of \(\boldsymbol{\varepsilon}\) and transform it to
\(\boldsymbol{e}\) to get the empirical estimates. For simplicity, in
this study, we replaced the variance-covariance matrix of residuals
\(\text{cov}(\boldsymbol{e}, \boldsymbol{e}) = \boldsymbol{R}\sigma^2\boldsymbol{R}' = \boldsymbol{R}\sigma^2\)
with a full-rank diagonal matrix
\(\text{diag}(\boldsymbol{R}\sigma^2)\), where \(\text{diag}(.)\) sets
the non-diagonal entries of a matrix to zeros. The resulting
distribution for \(\boldsymbol{Q}\) is
\(N(\boldsymbol{0}_n, \text{diag}(\boldsymbol{R}\sigma^2))\).

Distribution \(Q\) is derived from the correctly specified model.
However, if the model is misspecified, then the actual distribution of
residuals denoted as \(P\), will be different from \(Q\). For example,
if the data generating process contains variables correlated with any
column of \(\boldsymbol{X}\) but not included in \(\boldsymbol{X}\),
causing an omitted variable problem, \(P\) will be different from \(Q\)
because the residual operator obtained from the fitted regression model
will not be the same as \(\boldsymbol{R}\). Besides, if the
\(\boldsymbol{\varepsilon}\) follows a non-normal distribution such as a
multivariate log-normal distribution, \(P\) will usually be skewed and
has a long tail.

\subsubsection{\texorpdfstring{Kullback-Leibler divergence of \(P\) from
\(Q\)}{Kullback-Leibler divergence of P from Q}}\label{kullback-leibler-divergence-of-p-from-q}

Define a proper distance between distributions is usually easier than
define a proper distance between data plots. Given the actual residual
distribution \(Q\) and the reference residual distribution \(P\), we
used a distance measure based on Kullback-Leibler divergence
\citep{kullback1951information} to quantify the difference between two
distributions

@@

\noindent where \(p(.)\) is the probability density function for
distribution \(P\), and \(q(.)\) is the probability density function for
distribution \(Q\).

This distance measure was first proposed in \citet{li2023plot}. It was
mainly designed for measuring the effect size of non-linearity and
heteroskedasticity in a residual plot. \citet{li2023plot} have showed
that, for a classical normal linear regression model that omits a
necessary higher-order predictors \(\boldsymbol{Z}\), and incorrectly
assumes
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)\)
while in fact
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \boldsymbol{V})\)
with \(\boldsymbol{V}\) being an arbitrary symmetric positive
semi-definite matrix, \(Q\) can be represented as
\(N(\boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z, \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}))\).
Note that the variance-covariance matrix is replaced with the diagonal
matrix to ensure it is a full-rank matrix.

@@

\begin{align}
\label{eq:kl-2}
D_{KL} &= \frac{1}{2}\left(\log\frac{|\text{diag}(\boldsymbol{W})|}{|\text{diag}(\boldsymbol{R}\sigma^2)|} - n + \text{tr}(\text{diag}(\boldsymbol{W})^{-1}\text{diag}(\boldsymbol{R}\sigma^2)) + \boldsymbol{\mu}_z'(\text{diag}(\boldsymbol{W}))^{-1}\boldsymbol{\mu}_z\right),
\end{align}

\noindent where
\(\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z\),
and \(\boldsymbol{W} = \boldsymbol{R}\boldsymbol{V}\boldsymbol{R}\). The
assumed error variance \(\sigma^2\) is set to be
\(\text{tr}(\boldsymbol{V})/n\), which is the expectation of the
estimated variance.

\subsubsection{\texorpdfstring{Evaluation of Kullback-Leibler divergence
for non-normal
\(P\)}{Evaluation of Kullback-Leibler divergence for non-normal P}}\label{evaluation-of-kullback-leibler-divergence-for-non-normal-p}

For non-normal error \(\boldsymbol{\varepsilon}\), the actual residual
distribution \(P\) is unlikely to be a multivariate normal distribution.
Thus, equation \ref{eq:kl-2} given in \citet{li2023plot} will not be
applicable to models violating the normality assumption.

@@

\subsubsection{Approximation of the distance
measure}\label{approximation-of-the-distance-measure}

In the previous sections, we have defined a distance measure given in
equation \ref{eq:kl-0} for quantifying the difference between the actual
residual distribution \(P\) and an ideal reference distribution \(Q\).
You may have noticed that this distance measure can only be computed
when the data generating process is known. In reality, we often have no
knowledge about the data generating process, otherwise we do not need to
fit a linear regression model in the first place.

We tried to train a computer vision model to approximate this distance
measure with a residual plot. Let \(D\) be the result of equation
\ref{eq:kl-0}, and our estimator \(\hat{D}\) is formulated as

@@

\noindent where \(V_{h \times w}(.)\) is a plotting function that saves
a residuals vs fitted values plot with fixed aesthetic as an image with
\(h \times w\) pixels and three colour channels, \(f_{CV}(.)\) is a
computer vision model which takes an \(h \times w\) image as input and
predicts the distance in the domain \([0, +\infty)\).

With the approximated distance \(\hat{D}\), we will be able to know how
different the underlying distribution of the residuals is from a good
residual distribution. The approximated distance \(\hat{D}\) can also be
used as an index of the model violations. It also provides information
for the strength of the visual signal embedded in the residual plot.

The approximated distance \(\hat{D}\) is not expected to be the same as
the original distance \(D\). This is largely because information
contained in a single residual plot is limited and it may not be able to
summarise all the important characteristics of the residual
distribution. For a given residual distribution \(P\), we can generate
many different residual plots. Some of them share similar visual
patterns, but some of them could be visually very different from the
rest, especially for regression models with small \(n\). This suggests
the error of the approximation will vary depends on whether the observed
residual plot is representative or not.

\subsubsection{Model Violations Index
(MVI)}\label{sec:model-violations-index}

In Section \ref{output-formats}, we discuss the output formats of the
computer vision model presented in this paper. One notable advantage
highlighted is its interpretability. The approximated distance
\(\hat{D}\) is an estimator of the difference between the actual
residual distribution and the reference residual distribution. This
difference primarily arises from deviations in model assumptions. The
magnitude of \(D\) directly reflects the degree of these deviations,
thus making \(\hat{D}\) instrumental in forming a model violation index.

Note that the Kullback-Leibler divergence might be influenced by the
number of random variables involved in its evaluation. Generally, a
larger number of observations will lead to a greater distance \(D\).
However, this does not imply that \(\hat{D}\) fails to accurately
represent the extent of model violations. In fact, when examining
residual plots with more observations, we often observe a stronger
visual signal strength, as the underlying patterns are more likely to be
revealed, except in cases of significant overlapping.

@@

\subsection{Statistical testing based on the approximated
distance}\label{statistical-testing-based-on-the-approximated-distance}

\subsubsection{Lineup evaluation}\label{lineup-evaluation}

@@

To address this issue while adhering to the principle of visual
inference, we can compare the estimated distance \(\hat{D}\) to the
approximated distances for the null plots in a lineup. Specifically, if
a lineup comprises 20 plots, the null hypothesis \(H_0\) will be
rejected if \(\hat{D}\) exceeds the maximum approximated distance among
the \(m - 1\) null plots, denoted as
\(\max\limits_{1 \leq i \leq m-1} {\hat{D}_{null}^{(i)}}\), where
\(\hat{D}_{null}^{(i)}\) represents the approximated distance for the
\(i\)-th null plot. This approach is equivalent to the typical lineup
protocol requiring a 95\% significance level, where \(H_0\) is rejected
if the data plot is identified as the most distinct plot by the sole
participant. The approximated distance serves as a metric to quantify
the difference between the data plot and the null plots, as intended.

For lineups consisting of more than 20 plots, the 95\% significance
level can be maintained if the number of plots is a multiple of 20.
Specifically, for lineups comprising \(20p\) plots, where \(p\) is a
positive integer, we reject \(H_0\) if \(\hat{D}\) exceeds 95\%
\({\hat{D}_{null}^{(i)}}\) for \(i = 1, \ldots, 20p-1\). The p-value in
this case is given by
\(\frac{1}{m} + \frac{1}{m}\sum{i=1}^{m-1} I\left(\hat{D}_{null}^{(i)} > \hat{D}\right)\),
where \(I(\cdot)\) is the indicator function.

Moreover, if the number of plots in a lineup, denoted by \(m\), is
sufficiently large, the empirical distribution of
\({\hat{D}_{null}^{(i)}}\) can be viewed as an approximation of the null
distribution of the approximated distance. Consequently, quantiles of
the null distribution can be estimated using the sample quantiles
available in statistical software such as R, and these quantiles can be
utilized for decision-making purposes. The details of the sample
quantile computation can be found in \citet{hyndman1996sample}. For
instance, if \(\hat{D}\) is greater than or equal to the 95\% sample
quantile, denoted as \(Q_{null}(0.95)\), we can conclude that the
approximated distance for the actual residual plot is significantly
different from the approximated distance for null plots with a 95\%
significance level. Based on our experience, to obtain a stable estimate
of the 95\% quantile, the number of null plots, \(n_{null}\), typically
needs to be at least 100. However, if the null distribution exhibits a
long tail, a larger number of null plots may be required. Alternatively,
a p-value can be used to represent the probability of observing an event
equally or more extreme than the given event under the null hypothesis
\(H_0\), and it can be estimated by
\(\frac{1}{m}\sum_{i=1}^{m-1}I\left(\hat{D}_{null}^{(i)} \geq \hat{D}\right)\).

@@

\subsubsection{Bootstrapping the approximated
distance}\label{bootstrapping-the-approximated-distance}

@@

Similarly, bootstrap can be applied on the approximated distance
\(\hat{D}\). For each refitted model \(M_{boot}^{(i)}\), there will be
an associated residual plot \(V_{boot}^{(i)}\) which can be fed into the
computer vision model to obtain \(\hat{D}_{boot}^{(i)}\), where
\(i = 1,...,n_{boot}\), and \(n_{boot}\) is the number of bootstrapped
samples. If we are interested in the variation of \(\hat{D}\), we can
use \(\hat{D}_{boot}^{(i)}\) to estimate a confidence interval.

@@

\subsection{Generation of training
data}\label{generation-of-training-data}

\subsubsection{Data generating process}\label{data-generating-process}

@@

\subsubsection{Computation of
scagnostics}\label{computation-of-scagnostics}

@@

For each generated residual plot, we computed four scagnostics --
``Monotonic,'' ``Sparse,'' ``Splines,'' and ``Striped'' -- using the
\texttt{cassowaryr} R package \citep{mason2022cassowaryr}. These
computed measures, along with the number of observations from the fitted
model, were provided as the second input for the computer vision model.
Although other scagnostics are informative, they are currently
unavailable due to a fatal bug in the compiled C program of the
\texttt{interp} R package \citep{Albrecht2023interp}, which may
unpredictably crash the process. For reproducibility, we excluded these
scagnostics from the training data.

\subsubsection{Crafting a balanced training
set}\label{crafting-a-balanced-training-set}

@@

\subsection{Architecture of the computer vision
model}\label{architecture-of-the-computer-vision-model}

@@

\subsection{Training and hyperparameter
tuning}\label{training-and-hyperparameter-tuning}

@@

\subsection{Model evaluation methods}\label{model-evaluation-methods}

@@

\subsection{Optimized hyperparameters}\label{optimized-hyperparameters}

@@

\subsection{Demostration of MVI}\label{demostration-of-mvi}

As described in Section \ref{sec:model-violations-index} a Model
Violations Index (MVI) can be constructed using \(\hat{D}\). This
section provides guidelines and examples of MVI.

@@

In order to check the validity of the proposed computer vision model,
residual plots presented in the human subject experiment conducted by
\citet{li2023plot} will be assessed. This study has collected \(7974\)
human responses to \(1152\) lineups. Each lineup contains one randomly
placed actual residual plot and 19 null plots. Among the \(1152\)
lineups, \(24\) are attention check lineups in which the visual patterns
are designed to be extremely obvious and very different from the
corresponding to null plots, \(36\) are null lineups where all the
lineups consist of only null plots, \(279\) are lineups with uniform
predictor distribution evaluated by \(11\) participants, and the
remaining \(813\) are lineups with discrete, skewed or normal predictor
distribution evaluated by \(5\) participants. Attention check lineups
and null lineups will not be assessed in the following analysis.

@@

Figure \ref{fig:delta} indicates that the weighted detection rate
increases as the \(\delta\)-difference increases, particularly when the
\(\delta\)-difference is greater than zero. A negative
\(\delta\)-difference suggests that there is at least one null plot in
the lineup with a stronger visual signal than the actual residual plot.
In some instances, the weighted detection rate is close to one, yet the
\(\delta\)-difference is negative. This discrepancy implies that the
distance measure, or the approximated distance, may not perfectly
reflect actual human behaviour.

>>>>

\usepackage{lscape}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\def\tightlist{}
\usepackage{setspace}

@@

Generating a residual plot in most statistical software is often as
straightforward as executing a line of code or clicking a button.
However, accurately interpreting a residual plot can be challenging. A
residual plot can exhibit various visual features, but it is crucial to
recognize that some may arise from the characteristics of predictors and
the inherent randomness of the error, rather than indicating a violation
of model assumptions \citep{li2023plot}. Consider Figure
\ref{fig:false-finding} as an example, the residuals display a
triangular shape pointing to the left. While this might suggest
heteroskedasticity, it is important to avoid over-interpreting this
visual pattern. In this case, the fitted regression model is correctly
specified, and the triangular shape is actually a result of the skewed
distribution of the predictors, rather than indicating a flaw in the
model.

The concept of visual inference, as proposed by
\citet{buja2009statistical}, provides an inferential framework to assess
whether residual plots indeed contain visual patterns inconsistent with
the model assumptions. The fundamental idea involves testing whether the
true residual plot visually differs significantly from null plots, where
null plots are plotted with residuals generated from the residual
rotation distribution \citep{langsrud2005rotation}, which is a
distribution consistent with the null hypothesis \(H_0\) that the linear
regression model is correctly specified. Typically, the visual test is
accomplished through the lineup protocol, where the real residual plot
is embedded within a lineup alongside several null plots. If the real
residual plot can be distinguished from the lineup, it provides evidence
for rejecting \(H_0\).

@@

The utilization of computers to interpret data plots has a rich history,
with early efforts such as ``Scagnostics'' by \citet{tukey1985computer},
focusing on scatter plot diagnostics. \citet{wilkinson2005graph}
expanded on this work, introducing graph theoretic scagnostics, which
encompassed computable measures applied to planar proximity graphs.
These measures, including, but not limited to, ``Outlying'', ``Skinny'',
``Stringy'', ``Straight'', ``Monotonic'', ``Skewed'', ``Clumpy'', and
``Striated'' aimed to characterize outliers, shape, density, trend,
coherence and other characteristics of the data. While this approach has
been inspiring, there is a recognition \citep{buja2009statistical} that
it may not capture all the necessary visual features that differentiate
true residual plots from null plots. A more promising alternative
entails enabling machines to learn the function for extracting visual
features from residual plots. Essentially, this means empowering
computers to discern the crucial visual features for residual
diagnostics and determining the method to extract them.

@@

\subsection{Model Formulas}\label{model-formulas}

There are various configurations of the computer vision model that can
be used to assess residual plots. We discuss these configurations below
focusing on two key components of the model formula: the input and the
output format.

\subsubsection{Input Formats}\label{input-formats}

@@

Providing multiple residual plots to the model, such as a pair of plots,
a triplet or a lineup is also a possible option.
\citet{chopra2005learning} have shown that computer vision models
designed for image comparison can assess whether a pair of images are
similar or dissimilar. Applied to our specific problem, we can define
null plots of a fitted regression model to be similar to each other,
while considering true residual plots to be distinct from null plots of
any fitted regression model. A triplet constitutes a set of three
images, denoted as \(image_1\), \(image_2\) and \(image_3\). It is often
used to predict whether \(image_2\) or \(image_3\) is more similar to
\(image_1\), proving particularly useful for establishing rankings
between samples. For this setup, we can apply the same criteria to
define similarity between images. However, it is important to note that
these two approaches usually require additional considerations regarding
the loss function and, at times, non-standard training processes due to
shared weights between different convolutional blocks.

@@

We did not explore all the mentioned input formats due to the
significant costs tied to data preparation and model training. Taking
into account the implementation cost and the need for model
interpretability, we settled on the single residual plot input format.

\subsubsection{Output Formats}\label{output-formats}

@@

\subsection{Distance Between Residual
Plots}\label{distance-between-residual-plots}

@@

\subsubsection{Residual Distribution}\label{residual-distribution}

The distance measure proposed in this study takes into account the fact
that we tried to measure how different a residual plot is from a good
residual plot, or in other words, how different a given fitted
regression model is from a correctly specified model.

For a classical normal linear regression model, residuals
\(\boldsymbol{e}\) are derived from the fitted values
\(\hat{\boldsymbol{y}}\) and observed values \(\boldsymbol{y}\). Suppose
the data generating process is known and the regression model is
correctly specified, by the Frisch-Waugh-Lowell theorem
\citep{frisch1933partial}, residuals \(\boldsymbol{e}\) can also be
treated as random variables and written as a linear transformation of
the error \(\boldsymbol{\varepsilon}\) formulated as
\(\boldsymbol{e} = \boldsymbol{R}\boldsymbol{\varepsilon}\), where
\(\boldsymbol{R}=\boldsymbol{I}_n -\boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'\)
is the residual operator and an idempotent matrix, \(\boldsymbol{X}\) is
the design matrix, \(\boldsymbol{I}_n\) is a \(n\) by \(n\) identity
matrix, and \(n\) is the number of observations.

@@

Suppose the design matrix \(\boldsymbol{X}\) has linearly independent
columns, the trace of the hat matrix will equal to the number of columns
in \(\boldsymbol{X}\) denoted as \(k\). As a result, the rank of
\(\boldsymbol{R}\) is \(n - k\), and \(Q\) is a degenerate multivariate
distribution. To capture the characteristics of \(Q\), such as moments,
we can simulate a large numbers of \(\boldsymbol{\varepsilon}\) and
transform it to \(\boldsymbol{e}\) to get the empirical estimates. For
simplicity, in this study, we replaced the variance-covariance matrix of
residuals
\(\text{cov}(\boldsymbol{e}, \boldsymbol{e}) = \boldsymbol{R}\sigma^2\boldsymbol{R}' = \boldsymbol{R}\sigma^2\)
with a full-rank diagonal matrix
\(\text{diag}(\boldsymbol{R}\sigma^2)\), where \(\text{diag}(.)\) sets
the non-diagonal entries of a matrix to zeros. The resulting
distribution for \(\boldsymbol{Q}\) is
\(N(\boldsymbol{0}_n, \text{diag}(\boldsymbol{R}\sigma^2))\).

Distribution \(Q\) is derived from the correctly specified model.
However, if the model is misspecified, then the actual distribution of
residuals denoted as \(P\), will be different from \(Q\). For example,
if the data generating process contains variables correlated with any
column of \(\boldsymbol{X}\) but missing from \(\boldsymbol{X}\),
causing an omitted variable problem, \(P\) will be different from \(Q\)
because the residual operator obtained from the fitted regression model
will not be the same as \(\boldsymbol{R}\). Besides, if the
\(\boldsymbol{\varepsilon}\) follows a non-normal distribution such as a
multivariate log-normal distribution, \(P\) will usually be skewed and
has a long tail.

\subsubsection{\texorpdfstring{Distance of \(P\) from
\(Q\)}{Distance of P from Q}}\label{distance-of-p-from-q}

Define a proper distance between distributions is usually easier than
define a proper distance between data plots. Given the true residual
distribution \(Q\) and the reference residual distribution \(P\), we
used a distance measure based on Kullback-Leibler divergence
\citep{kullback1951information} to quantify the difference between two
distributions

@@

\noindent where \(p(.)\) and \(q(.)\) are the probability density
functions for distribution \(P\) and distribution \(Q\) respectively.

This distance measure was first proposed in \citet{li2023plot}. It was
mainly designed for measuring the effect size of non-linearity and
heteroskedasticity in a residual plot. \citet{li2023plot} have derived
that, for a classical normal linear regression model that omits a
necessary higher-order predictors \(\boldsymbol{Z}\), and incorrectly
assumes
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)\)
while in fact
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \boldsymbol{V})\)
with \(\boldsymbol{V}\) being an arbitrary symmetric positive
semi-definite matrix, \(Q\) can be represented as
\(N(\boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z, \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}))\).
Note that the variance-covariance matrix is replaced with the diagonal
matrix to ensure it is a full-rank matrix.

@@

\begin{align}
\label{eq:kl-2}
D_{KL} &= \frac{1}{2}\left(\log\frac{|\text{diag}(\boldsymbol{W})|}{|\text{diag}(\boldsymbol{R}\sigma^2)|} - n + tr(\text{diag}(\boldsymbol{W})^{-1}\text{diag}(\boldsymbol{R}\sigma^2)) + \boldsymbol{\mu}_z'(\text{diag}(\boldsymbol{W}))^{-1}\boldsymbol{\mu}_z\right),
\end{align}

\noindent where
\(\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z\),
and \(\boldsymbol{W} = \boldsymbol{R}\boldsymbol{V}\boldsymbol{R}\). The
assumed error variance \(\sigma^2\) is set to be
\(tr(\boldsymbol{V})/n\), which is the expectation of the estimated
variance.

\subsubsection{\texorpdfstring{Non-normal
\(P\)}{Non-normal P}}\label{non-normal-p}

For non-normal error \(\boldsymbol{\varepsilon}\), the true residual
distribution \(P\) is unlikely to be a multivariate normal distribution.
Thus, equation \ref{eq:kl-2} given in \citet{li2023plot} will not be
applicable to models violating the normality assumption.

@@

\subsection{Distance Estimation}\label{distance-estimation}

In the previous sections, we have defined a distance measure given in
equation \ref{eq:kl-0} for quantifying the difference between the true
residual distribution \(P\) and an ideal reference distribution \(Q\).
You may have noticed that this distance measure can only be computed
when the data generating process is known. In reality, we often have no
knowledge about the data generating process, otherwise we do not need to
fit a linear regression model in the first place.

We tried to train a computer vision model to estimate this distance
measure with a residual plot. Let \(D\) be the result of equation
\ref{eq:kl-0}, and our estimator \(\hat{D}\) is formulated as

@@

\noindent where \(V_{h \times w}(.)\) is a plotting function that saves
a residuals vs fitted values plot with fixed aesthetic as an RGB image
with \(h \times w\) pixels, \(f_{CV}(.)\) is a computer vision model
which takes an \(h \times w\) image as input and predicts the distance
in the domain \([0, +\infty)\).

With the estimated distance \(\hat{D}\), we will be able to know how
different the underlying distribution of the residuals is from a good
residual distribution. \(\hat{D}\) can also be used as an index of the
model violations indicating the strength of the visual signal embedded
in the residual plot.

It is not expected that \(\hat{D}\) will be equal to original distance
\(D\). This is largely because information contained in a single
residual plot is limited and it may not be able to summarise all the
important characteristics of the residual distribution. For a given
residual distribution \(P\), many different residual plots can be
simulated, where many will share similar visual patterns, but some of
them could be visually very different from the rest, especially for
regression models with small \(n\). This suggests the error of the
estimation will vary depends on whether the input residual plot is
representative or not.

\subsection{Model Violations Index
(MVI)}\label{sec:model-violations-index}

\(\hat{D}\) is an estimator of the difference between the true residual
distribution and the reference residual distribution. This difference
primarily arises from deviations in model assumptions. The magnitude of
\(D\) directly reflects the degree of these deviations, thus making
\(\hat{D}\) instrumental in forming a model violations index.

Note that Equation \ref{eq:kl-0} might be influenced by the number of
random variables involved in its evaluation. Generally, a larger number
of observations will lead to a greater distance \(D\). However, this
does not imply that \(\hat{D}\) fails to accurately represent the extent
of model violations. In fact, when examining residual plots with more
observations, we often observe a stronger visual signal strength, as the
underlying patterns are more likely to be revealed, except in cases of
significant overlapping.

@@

\subsection{Statistical testing}\label{statistical-testing}

\subsubsection{Lineup Evaluation}\label{lineup-evaluation}

@@

To address this issue while adhering to the principle of visual
inference, we can compare the estimated distance \(\hat{D}\) to the
estimated distances for the null plots in a lineup. Specifically, if a
lineup comprises 20 plots, the null hypothesis \(H_0\) will be rejected
if \(\hat{D}\) exceeds the maximum estimated distance among the
\(m - 1\) null plots, denoted as
\(\max\limits_{1 \leq i \leq m-1} {\hat{D}_{null}^{(i)}}\), where
\(\hat{D}_{null}^{(i)}\) represents the estimated distance for the
\(i\)-th null plot. This approach is equivalent to the typical lineup
protocol requiring a 95\% significance level, where \(H_0\) is rejected
if the data plot is identified as the most distinct plot by the sole
participant. The estimated distance serves as a metric to quantify the
difference between the data plot and the null plots, as intended.

For lineups consisting of more than 20 plots, the 95\% significance
level can be maintained if the number of plots is a multiple of 20.
Specifically, for lineups comprising \(20t\) plots, where \(l\) is a
positive integer, we reject \(H_0\) if \(\hat{D}\) exceeds 95\%
\({\hat{D}_{null}^{(i)}}\) for \(i = 1, \ldots, 20t-1\). The \(p\)-value
in this case is given by
\(\frac{1}{20t} + \frac{1}{20t}\sum{i=1}^{20t-1} I\left(\hat{D}_{null}^{(i)} > \hat{D}\right)\),
where \(I(\cdot)\) is the indicator function.

Moreover, if the number of plots in a lineup, denoted by \(m\), is
sufficiently large, the empirical distribution of
\({\hat{D}_{null}^{(i)}}\) can be viewed as an approximation of the null
distribution of the approximated distance. Consequently, quantiles of
the null distribution can be estimated using the sample quantiles
available in statistical software such as R, and these quantiles can be
utilized for decision-making purposes. The details of the sample
quantile computation can be found in \citet{hyndman1996sample}. For
instance, if \(\hat{D}\) is greater than or equal to the 95\% sample
quantile, denoted as \(Q_{null}(0.95)\), we can conclude that the
estimated distance for the true residual plot is significantly different
from the estimated distance for null plots with a 95\% significance
level. Based on our experience, to obtain a stable estimate of the 95\%
quantile, the number of null plots, \(n_{null}\), typically needs to be
at least 100. However, if the null distribution exhibits a long tail, a
larger number of null plots may be required. Alternatively, a
\(p\)-value can be used to represent the probability of observing an
event equally or more extreme than the given event under the null
hypothesis \(H_0\), and it can be estimated by
\(\frac{1}{m}\sum_{i=1}^{m-1}I\left(\hat{D}_{null}^{(i)} \geq \hat{D}\right)\).

@@

\subsubsection{Bootstrapping}\label{bootstrapping}

@@

Similarly, bootstrap can be applied on the estimated distance
\(\hat{D}\). For each refitted model \(M_{boot}^{(i)}\), there will be
an associated residual plot \(V_{boot}^{(i)}\) which can be fed into the
computer vision model to obtain \(\hat{D}_{boot}^{(i)}\), where
\(i = 1,...,n_{boot}\), and \(n_{boot}\) is the number of bootstrapped
samples. If we are interested in the variation of \(\hat{D}\), we can
use \(\hat{D}_{boot}^{(i)}\) to estimate a confidence interval.

@@

\subsection{Data Genearation}\label{data-genearation}

\subsubsection{Simulation Scheme}\label{simulation-scheme}

@@

\subsubsection{Scagnostics}\label{scagnostics}

@@

For each generated residual plot, we computed four scagnostics --
``Monotonic,'' ``Sparse,'' ``Splines,'' and ``Striped'' -- using the
\texttt{cassowaryr} R package \citep{mason2022cassowaryr}. These
computed measures, along with the number of observations from the fitted
model, were provided as the second input for the computer vision model.
While other scagnostics provide valuable insights, they come with high
computational costs and are not suitable for quick inference.

\subsubsection{Balanced Dataset}\label{balanced-dataset}

@@

\subsection{Model Architecture}\label{model-architecture}

{\centering \includegraphics[width=1\linewidth]{cnn}

\caption{Diagram of the architecture of the optimized computer vision model.}\label{fig:unnamed-chunk-2}
\end{figure}

@@

\subsection{Model Training}\label{model-training}

@@

\subsection{Evaluation Methods}\label{evaluation-methods}

@@

understand what the convolutional layer is actually doing (read vgg16
doc)

@@

@@

In order to check the validity of the proposed computer vision model,
residual plots presented in the human subject experiment conducted by
\citet{li2023plot} will be assessed. This study has collected \(7974\)
human responses to \(1152\) lineups. Each lineup contains one randomly
placed true residual plot and 19 null plots. Among the \(1152\) lineups,
\(24\) are attention check lineups in which the visual patterns are
designed to be extremely obvious and very different from the
corresponding to null plots, \(36\) are null lineups where all the
lineups consist of only null plots, \(279\) are lineups with uniform
predictor distribution evaluated by \(11\) participants, and the
remaining \(813\) are lineups with discrete, skewed or normal predictor
distribution evaluated by \(5\) participants. Attention check lineups
and null lineups will not be assessed in the following analysis.

@@

Figure \ref{fig:delta} indicates that the weighted detection rate
increases as the \(\delta\)-difference increases, particularly when the
\(\delta\)-difference is greater than zero. A negative
\(\delta\)-difference suggests that there is at least one null plot in
the lineup with a stronger visual signal than the true residual plot. In
some instances, the weighted detection rate is close to one, yet the
\(\delta\)-difference is negative. This discrepancy implies that the
distance measure, or the approximated distance, may not perfectly
reflect actual human behaviour.
